\toolnameend: A Tool for Visualizing Patterns of Spreadsheet Function Combinations

### 2016-04-26 doc/paper.tex
More paper work. Fixes #122, fixes #132, fixes #133, fixes #140, fixes #141, fixes #145, fixes #146.

	\section{Introduction} Spreadsheets surround us. They serve key roles in
	industry, where as much as 95\% of U.S. financial firms use them daily
	\cite{panko2008sarbanes}, and academia, where teachers can use them to track
	research and students' grades, not to mention many other applications
	\cite{ko2011state}. It's this flexibility that grants them their allure: while
	a novice can still work without much experience, spreadsheets offer hundreds of
	specialized functions to fulfill a wide range of advanced needs, from complex
	statistics to text manipulation \cite{nardi1990spreadsheet}. As such, it should
	be no surprise that tens of millions of workers rely on these programs daily \cite{scaffidi2005estimating}.

	However, this ubiquity is not without danger; as more people adopt these tools
	and make larger and more complex sheets, the cost of failure grows too. To illustrate this, the European Spreadsheet Risks Interest Group (EuSpRIG) documents
	many horror stories of spreadsheets gone 
	wrong\footnote{http://www.eusprig.org/horror-stories.htm}. One story tells
	of a 2013 economics paper which drew connections between debt and
	national growth. Its findings shaped many political initiatives throughout the
	United States and Europe. However, other researchers soon discovered that a
	critical selection error in the research spreadsheets hampered the results and
	that, by correcting this error, the findings reversed completely! And this
	is only one of several accounts; even if most spreadsheet errors are relatively
	benign, poor spreadsheet practice can risk millions, if not billions, of
	dollars \cite{powell2009impact}.

	From these stories, we can see the dire importance of understanding how people
	actually use spreadsheets, if only to avoid disasters. Fortunately, spreadsheet
	researchers have supported this by amassing spreadsheet collections from
	diverse sources, industrial to academic. Collections like EUSES
	\cite{fisher2005euses}, Fuse \cite{barik2015fuse}, and the Enron corpus
	\cite{hermans2015enron} have already fueled productive research across the
	field, such as in Hermans' \cite{hermans2012detecting} and Jansen's
	\cite{jansen2015code} search for code smells in spreadsheet environments or
	Aivaloglou and colleague's work on making a grammar to capture every possible
	spreadsheet program in Excel \cite{aivaloglou2015grammar}. Additionally, some
	research has focused on comparing these collections, as Jansen did in
	2015, to discover variation in spreadsheet techniques over time and location
	\cite{jansen2015enron}.

	Considering the value of spreadsheets and all of these resources to
	describe their use, it becomes necessary that we design tools to make sense of
	it all. Tools, then, that empower people to explore datasets are rife with the
	potential to improve the current design of spreadsheets themselves and educate
	new generations of spreadsheets users on how to work best.

@@

	\section{Related Work} \label{related-work} Visualizations have long been prized for their ability to
	communicate information quickly and efficient \cite{baeza1999modern}; the
	visualization of spreadsheets, therefore, has offered much room for research.
	Several previous tools work within a spreadsheet, seeking to clarify the
	connections between the visible values in cells, the formulae that produce
	them, and the dependencies between cells that inform them. Igarashi and
	colleague's fluid visualizations \cite{igarashi1998fluid}, for example, do this
	by imposing lines, color, and animation over the spreadsheet to visually
	explain these tangled connections. Likewise, Clermont explored ways of grouping
	cells by color and border that were related through similar functions,
	neighbors, or references \cite{clermont2003scalable}, and Hipfl extended this
	work by incorporating the layouts and labels of spreadsheets
	\cite{hipfl2008using}. Other tools focus on creating visualizations external to
	the sheets. Hermans and colleagues address a spreadsheet programmer's
	information needs by making dataflow diagrams from individual sheets through the
	tools GyroSAT \cite{hermans2011supporting} and Breviz \cite{hermans2011breviz}.
	\toolname is among the latter class, creating standalone visualizations about
	spreadsheets. In contrast to Hermans' work, our work focuses on
	representing functions and how they comprise complex formula, rather than
	exploring the data structures of any particular spreadsheet.  \par

	Many other tools inspect the structure of formulae but often from the
	perspective of finding bad design.  Following Fowler's initial description of
	code smells in object-oriented design \cite{fowler2009refactoring}, many
	researchers have applied similar observations to spreadsheets and have
	cataloged classes of problematic formulae \cite{hermans2012detecting}
	\cite{cunha2012towards} \cite{asavametha2012detecting}, often accompanying them
	with detection tools \cite{abreu2014smelling} and investigations
	\cite{jansen2015code}. For example, Hermans, leveraging her work with the
	aforementioned dataflow diagrams, has worked with assessing common
	inter-worksheet smells, such as feature envy and inappropriate intimacy,
	wherein references between worksheets (rather than cells) in a file suggest a
	problem \cite{hermans2012detectinginter}. Other tools, like Badam's Excel
	refactoring \cite{badame2012refactoring}, are concerned more with understanding
	the formula in service of changing them. However, these studies define
	standards of poor design and orient their searches around them. Though
	\toolname supports the search for bad design, as we will see, the tool itself
	takes a much more agnostic approach in presentation, offering examples without
	regard for design quality. \par

	Implicit in this study is an exploration of the Excel API and how people use
	it. Previous research, such as the first study on the Enron spreadsheets
	\cite{hermans2015enron} or in its comparison with the EUSUS corpus
	\cite{jansen2015enron}, quantify how ofter users employ certain functions but
	not how they combine them. Researchers have applied such questions to other
	contexts as well, such as Murphy and colleague's study into how Java developers
	use Eclipse and what commands they use most often \cite{murphy2006java}. Others
	focus on the documentation of the API itself: Robillard and DeLine, for example
	conducted a series of surveys and interviews to diagnose common problems with
	learning APIs, and they found that code examples, one of the focuses of this
	tool, is one of the most important aids to have in learning a new system
	\cite{robillard2011field}. Though this paper applies the tool to the Enron
	corpus specifically, the tool itself is not limited to this; it can be readily
	applied to any collection of Excel spreadsheets a user gives it.

@@

	\subsection{Design Goals} \label{goals} In visualizing the spreadsheet data, we
	outlined the core goals for what the tool should accomplish. Furthermore, with
	these goals, we also considered the likely challenges of achieving those goals
	and particular outcomes which we considered beyond the scope of the research.
	We present both sides of these goals below:

@@

		\item [1] \textbf{Provide an interactive environment for free exploration.}
		There are a lot of data in some of these spreadsheet corpora, and the user should be
		able to pick for themselves which to explore while ignoring the rest.
		Furthermore, this exploration should not be designed around answering one
		specific question but rather be open-ended.

		\item [!1] \textbf{Do not make judgments about good and bad design.} Though
		users may certainly approach the tool with the intention of finding particular
		formulae, \toolname will not prioritize any on metrics of good or bad
		practice.

		\item [2] \textbf{Emphasize the quantitative patterns in formula
			construction.} \toolname should address the questions of how often the
		spreadsheet programmers used a certain function and where they used it. In
		this way, it must show metrics, such as frequency of use and depth of
		function nesting, of the dataset it conveys.

		\item [!2] \textbf{Do not overwhelm with numbers.} For popular functions
		especially, we anticipated some particularly dense trees. Though everything
		should be accessible in accordance with Goal 1, we should design in a way that
		reduces noise.

@@

	\subsection{Walkthrough of the Tool} \label{sec:walkthrough}Before discussing
	the minutiae of tool's development, though, we will give a basic example of how
	to use the tool. For this, we will answer this question: \textit{ What kinds of
		functions do people use to define the condition in an IF function?} The full
	visualization that answers this is available in Figure~\ref{fig:fullpic};
	however, to understand the interaction, we will start from the beginning. \par

>>>>

	\section{Introduction} Spreadsheets surround us. They serve key roles in
	industry, where as much as 95\% of U.S. financial firms use them daily~\cite{panko2008sarbanes}, and academia, where teachers can use them to track
	research and students' grades, not to mention many other applications~\cite{ko2011state}. It's this flexibility that grants them their allure: while
	a novice can still work without much experience, spreadsheets offer hundreds of
	specialized functions to fulfill a wide range of advanced needs, from complex
	statistics to text manipulation~\cite{nardi1990spreadsheet}. As such, it should
	be no surprise that tens of millions of workers rely on these programs daily~\cite{scaffidi2005estimating}.

	However, this ubiquity is not without danger; as more people adopt these tools
	and make larger and more complex sheets, the cost of failure grows too. To illustrate this, the European Spreadsheet Risks Interest Group (EuSpRIG) documents
	many horror stories of spreadsheets gone 
	wrong\footnote{http://www.eusprig.org/horror-stories.htm}. One story tells
	of a 2013 economics paper which drew connections between debt and
	national growth. Its findings shaped many political initiatives throughout the
	United States and Europe. However, other researchers soon discovered that a
	critical selection error in the research spreadsheets hampered the results and
	that, by correcting this error, the findings reversed completely! And this
	is only one of several accounts; even if most spreadsheet errors are relatively
	benign, poor spreadsheet practice can risk millions, if not billions, of
	dollars~\cite{powell2009impact}.

	From these stories, we can see the dire importance of understanding how people
	actually use spreadsheets. Fortunately, spreadsheet
	researchers have supported this by amassing spreadsheet collections from
	diverse sources, industrial to academic. Collections like EUSES~\cite{fisher2005euses}, 
	Fuse~\cite{barik2015fuse}, and the Enron corpus~\cite{hermans2015enron} have already fueled productive research across the
	field, such as in Hermans'~\cite{hermans2012detecting} and Jansen's~\cite{jansen2015code} search for code smells in spreadsheet environments or
	Aivaloglou and colleague's work on making a grammar to capture every possible
	spreadsheet program in Excel~\cite{aivaloglou2015grammar}. Additionally, some
	research has focused on comparing these collections, as Jansen did in
	2015, to discover variation in spreadsheet techniques over time and location~\cite{jansen2015enron}.

	Considering the value of spreadsheets and all of these resources to
	describe their use, it becomes necessary that we design tools to better enable the
	exploration of these datasets and simplify the complex concepts that occur in
	spreadsheet design. One such concept is that of formula construction; functions
	are an important part of using spreadsheets efficiently, yet the number of their potential
	combinations quickly becomes massive when you consider that there are hundreds of them.
	Tools, then, that empower people to explore questions about which functions are used together 
	in practice have the potential to improve the current design of spreadsheets themselves and educate
	new generations of spreadsheets users on how to work best.

@@

	\section{Related Work} \label{related-work} Visualizations excel for their ability to
	communicate information quickly and efficiently~\cite{baeza1999modern}; the
	visualization of spreadsheets, therefore, has offered much room for research.
	Several previous tools work within a spreadsheet, seeking to clarify the
	connections between the visible values in cells, the formulae that produce
	them, and the dependencies between cells that inform them. Igarashi and
	colleague's fluid visualizations~\cite{igarashi1998fluid}, for example, do this
	by imposing lines, color, and animation over the spreadsheet to visually
	explain these tangled connections. Likewise, Clermont explored ways of grouping
	cells by color and border that were related through similar functions,
	neighbors, or references~\cite{clermont2003scalable}, and Hipfl extended this
	work by incorporating the layouts and labels of spreadsheets~\cite{hipfl2008using}. Other tools focus on creating visualizations external to
	the sheets. Hermans and colleagues address a spreadsheet programmer's
	information needs by making dataflow diagrams from individual sheets through the
	tools GyroSAT~\cite{hermans2011supporting} and Breviz~\cite{hermans2011breviz}.
	\toolname is among the latter class, creating standalone visualizations about
	spreadsheets instead of drawing connections inside the spreadsheet itself. In contrast to Hermans' work, our work focuses on
	representing functions and how they comprise complex formula, rather than
	exploring the data structures of any particular spreadsheet.  \par

	Many other tools inspect the structure of formulae but often from the
	perspective of finding bad design.  Following Fowler's initial description of
	code smells in object-oriented design~\cite{fowler2009refactoring}, many
	researchers have applied similar observations to spreadsheets and have
	cataloged classes of problematic formulae~\cite{hermans2012detecting}~\cite{cunha2012towards}~\cite{asavametha2012detecting}, often accompanying them
	with detection tools~\cite{abreu2014smelling} and investigations into real datasets~\cite{jansen2015code}. For example, Hermans, leveraging her work with the
	aforementioned dataflow diagrams, has worked with assessing common
	inter-worksheet smells, such as feature envy and inappropriate intimacy,
	wherein references between worksheets (rather than cells) in a file suggest a
	problem~\cite{hermans2012detectinginter}. Other tools, like Badam's Excel
	refactoring~\cite{badame2012refactoring}, are concerned more with understanding
	the formula in service of changing them. However, these studies define
	standards of poor design and orient their searches around them. Though
	\toolname supports the search for bad design, as we will see, the tool itself
	takes a much more agnostic approach in presentation, offering examples without
	regard for design quality. \par

	Implicit in this study is an exploration of the Excel API and how people use
	it. Previous research, such as the first study on the Enron spreadsheets~\cite{hermans2015enron} or in its comparison with the EUSUS corpus~\cite{jansen2015enron}, quantify how ofter users employ certain functions but
	not how they combine them. Researchers have applied such questions to other
	contexts as well, such as Murphy and colleague's study into how Java developers
	use Eclipse and what commands they use most often~\cite{murphy2006java}. Others
	focus on the documentation of the API itself: Robillard and DeLine, for example
	conducted a series of surveys and interviews to diagnose common problems with
	learning APIs, and they found that code examples, one of the focuses of this
	tool, is one of the most important aids to have in learning a new system~\cite{robillard2011field}. Though this paper applies the tool to the Enron
	corpus specifically, the tool itself is not limited to this; it can be readily
	applied to any collection of Excel spreadsheets a user gives it.

@@

	\subsection{Design Goals} \label{goals} In visualizing the spreadsheet data, we
	follow a few core goals for what the tool should accomplish. Furthermore, with
	these goals, we also considered the likely challenges of achieving those goals
	and particular outcomes which we considered beyond the scope of the research.
	We present both sides of these goals below:

@@

		\item [1] \textbf{Provide an interactive environment for free exploration.}
		There are a lot of data in some of these spreadsheet corpora, and the user should be
		able to pick for themselves which to explore while ignoring the rest.
		Furthermore, this exploration should not be designed around answering one
		specific question but rather helping to answer a variety of questions.

		\item [!1] \textbf{Do not make judgments about good and bad design.} Though
		users may certainly approach the tool with the intention of exploring particular
		formulae, \toolname will not prioritize any on metrics of good or bad
		practice.

		\item [2] \textbf{Emphasize the quantitative patterns in formula
			construction.} \toolname should address the questions of how often the
		spreadsheet programmers used a certain function and where they used it. To do
		this, it must show metrics, such as frequency of use and depth of
		function nesting, of the dataset it conveys.

		\item [!2] \textbf{Do not overwhelm with numbers.} For popular functions
		especially, we anticipated that we'd have to show a lot of related data
		at once. Though everything
		should be accessible in accordance with Goal 1, we should design in a way that
		reduces noise.

@@

	\subsection{Walkthrough of the Tool} \label{sec:walkthrough}Before discussing
	the minutiae of tool's development, we will give a basic example of how
	to use the tool. For this, we will answer this question: \textit{ What kinds of
		functions do people use to define the condition in an IF function?} The full
	visualization that answers this is available in Figure~\ref{fig:fullpic};
	however, to understand the interaction, we will start from the beginning. \par
